---
title: "Decoupling or Illusion? Analysing Emission Intensity, Economic Growth, and Sectoral Energy Transitions in India (2000–2022)"
author: "Avyay Toprani"
date: today
---

# Introduction

Economic growth and carbon emissions have historically moved together. 
As countries industrialise, they consume more energy, burn more fossil 
fuels, and emit more greenhouse gases. The central challenge of 
sustainable development is breaking this relationship, growing 
economically while reducing or stabilising emissions.

India presents a particularly compelling case for this analysis. As the 
world's fifth largest economy and third largest emitter of CO2 during 
the study period, India's development trajectory carries significant 
implications for global climate outcomes. India has committed under its 
Nationally Determined Contributions (NDCs) to reduce the emissions 
intensity of its GDP by 45% by 2030 relative to 2005 levels, an 
ambitious target that assumes a fundamental decoupling of economic 
activity from carbon output.

Yet the nature of this decoupling remains contested. India's economy 
underwent a structural shift toward services during 2000 to 2022, a 
sector significantly less energy intensive than manufacturing or 
agriculture. This raises a critical question: is the observed reduction 
in emission intensity a genuine result of cleaner energy and greater 
efficiency, or is it a statistical artifact of economic restructuring?

This study investigates this question rigorously using three 
complementary analytical frameworks. First, the Tapio decoupling index 
is applied to classify the year by year relationship between GDP growth 
and CO2 emissions from 2000 to 2022. Second, the Log Mean Divisia 
Index (LMDI) decomposition method is used to isolate the specific 
drivers of emission changes, separating the effects of economic 
activity, sectoral structure, energy intensity, and energy mix. Third, 
the Environmental Kuznets Curve (EKC) hypothesis is tested to assess 
whether India follows the theoretical inverted U relationship between 
income and emissions, and to estimate whether a peak emissions point 
is approaching.

Together these analyses provide a comprehensive and evidence based 
assessment of whether India's growth was genuinely decoupling from its 
carbon footprint during the study period, or whether the appearance of 
progress masks deeper structural challenges.

# Literature Review

## The Decoupling Debate in Developing Economies

The relationship between economic growth and carbon emissions has been 
studied extensively across both developed and developing economies. The 
foundational concern is straightforward: if emissions grow proportionally 
with GDP, then economic development and climate targets become 
structurally incompatible. The concept of decoupling, where emissions 
grow slower than or independently of GDP, has therefore become central 
to sustainability research.

Early decoupling studies relied on the OECD decoupling factor, a simple 
ratio of emission growth to GDP growth. However, this measure proved 
highly sensitive to the choice of base and end years, producing 
inconsistent results across studies. Tapio (2005) addressed this 
limitation by introducing an elasticity-based decoupling index that 
classifies the GDP-emissions relationship into eight distinct states, 
offering a more stable and nuanced year by year assessment. This 
framework has since become widely adopted in the literature.

## Existing Studies on India

Several studies have applied decoupling and decomposition frameworks to 
India, though with important limitations that this study addresses. A 
comparative analysis of India, Brazil and China for the period 2000 to 
2018 using the Tapio model found that India's decoupling status showed 
continuous fluctuations, in contrast to China which moved toward strong 
decoupling after 2012. India predominantly experienced weak decoupling 
and expansive coupling during this period, suggesting that emissions 
continued to grow alongside the economy rather than separating from it.

Studies applying the LMDI decomposition method to India have largely 
focused on specific sectors rather than the economy as a whole. Research 
on India's transportation sector from 2001 to 2020 found that economic 
advancement and population scale were the primary drivers of rising CO2 
emissions, while energy efficiency improvements partially offset these 
increases. Similarly, studies on India's manufacturing sector have used 
combined Tapio and LMDI approaches but remain limited to organised 
manufacturing and do not account for the broader structural shift of 
India's economy toward services during this period.

On the EKC hypothesis, empirical evidence for India is mixed. A study 
covering 1991 to 2018 confirmed the existence of an inverted U-shaped 
EKC relationship for India, but found that CO2 levels continued rising 
past the theoretical turnaround point, suggesting that the income 
threshold alone does not determine emission trajectories. Research by 
Sinha and Shahbaz (2018) similarly found evidence of an inverted U-shaped 
EKC for India with a turnaround point at approximately USD 2,937 per 
capita, while also demonstrating that renewable energy generation 
played a significant role in moderating emissions.

## The Gap This Study Fills

Despite this body of work, three important gaps remain. First, most 
existing studies for India either end before 2020 or do not extend to 
2022, missing the post-pandemic recovery period which saw significant 
shifts in both GDP growth and energy demand. Second, no study to our 
knowledge applies all three frameworks, Tapio decoupling, LMDI 
decomposition, and EKC regression, together in an integrated analysis 
for India's economy as a whole over the 2000 to 2022 period. Third, the 
role of economic structural change, specifically the expansion of the 
services sector, as a confounding factor in India's apparent emission 
intensity improvements has not been rigorously isolated using 
decomposition methods at the national level.

This study addresses all three gaps by applying a unified analytical 
framework to the full 2000 to 2022 period, using the most recent 
available data from EDGAR, the World Bank, and the Central Electricity 
Authority of India.

# Methodology

## The Tapio Decoupling Index

The Tapio decoupling index, introduced by Tapio (2005), measures the 
elasticity of CO2 emissions with respect to GDP growth on a year by 
year basis. Unlike earlier OECD decoupling measures, the Tapio index 
is not sensitive to the choice of base year, making it more robust for 
longitudinal analysis. The index is calculated as follows:

$$e = \frac{\%\Delta CO_2}{\%\Delta GDP} = \frac{(CO_{2,t} - CO_{2,t-1})/CO_{2,t-1}}{(GDP_t - GDP_{t-1})/GDP_{t-1}}$$

Where $e$ is the Tapio decoupling index, $CO_{2,t}$ is total CO2 
emissions in year $t$, and $GDP_t$ is real GDP in year $t$. Based on 
the value of $e$, each year is classified into one of eight decoupling 
states. The three most relevant states for this study are strong 
decoupling ($e < 0$, GDP growing), weak decoupling ($0 < e < 0.8$, 
GDP growing), and expansive negative coupling ($e > 1.2$, GDP growing), 
representing the best, moderate, and worst case scenarios respectively 
for India's emissions trajectory.

**Data Processing:**

The Tapio index is calculated using India's total national CO2 
emissions from the EDGAR TOTALS BY COUNTRY dataset and real GDP in 
constant 2015 USD from the World Bank World Development Indicators. 
Both series were filtered to the study period 2000 to 2022 and merged 
into a unified panel dataset. Year on year percentage changes were 
computed for both variables, with 2000 serving as the base year. The 
index was calculated for each year from 2001 to 2022, producing 22 
annual observations. Each observation was then classified into one of 
Tapio's eight decoupling states based on the threshold values defined 
in the original framework.

## LMDI Decomposition

The Log Mean Divisia Index (LMDI) decomposition method, developed by 
Ang (2004), is used to identify the specific drivers of changes in 
India's CO2 emissions over the study period. LMDI is preferred over 
earlier decomposition methods because it produces an exact decomposition 
with no unexplained residual term.

Total CO2 emissions are expressed as the product of four factors across 
sectors:

$$C = \sum_{i} GDP \times \frac{GDP_i}{GDP} \times \frac{E_i}{GDP_i} \times \frac{C_i}{E_i}$$

Where $GDP$ is total real output, $GDP_i/GDP$ is the share of sector 
$i$ in total output representing the structure effect, $E_i/GDP_i$ is 
energy intensity of sector $i$ representing the intensity effect, and 
$C_i/E_i$ is the emission factor of sector $i$ representing the 
emission factor effect. 

While the full LMDI framework decomposes emission changes into four 
effects, namely the activity effect, structure effect, energy intensity 
effect, and emission factor effect, this study combines the final two 
into a single composite intensity effect. This approach is adopted due 
to the unavailability of consistent sectoral energy consumption data 
for India at the required granularity across the full study period. 
The combined intensity effect captures both changes in energy use per 
unit of output and changes in the carbon content of energy, and is a 
well established approach in the LMDI literature when sectoral energy 
data are unavailable. The decomposition identity used in this study is 
therefore:

$$\Delta C = \Delta C_{activity} + \Delta C_{structure} + \Delta C_{intensity}$$

Where $\Delta C_{intensity}$ represents the combined effect of energy 
intensity and emission factor changes across sectors.

Each component is calculated using the LMDI additive formula:

$$\Delta V = \sum_{i} L(V_{i}^{T}, V_{i}^{0}) \ln \left(\frac{x_{i,k}^{T}}{x_{i,k}^{0}}\right)$$

Where $L(V_{i}^{T}, V_{i}^{0})$ is the logarithmic mean of emissions 
in the base and target years, defined as:

$$L(A, B) = \frac{A - B}{\ln A - \ln B}$$

**Data Processing:**

The EDGAR IPCC 2006 dataset provides CO2 emissions for India across 
22 individual source categories for each year of the study period. 
To align with the sectoral structure required for LMDI decomposition 
and to ensure interpretability of results, these 22 categories were 
aggregated into seven broader sector groups as follows:

- **Power** — Main Activity Electricity and Heat Production
- **Industry** — Manufacturing Industries and Construction, Petroleum 
Refining, Cement Production, Lime Production, Glass Production, 
Other Process Uses of Carbonates, Chemical Industry, Metal Industry
- **Transport** — Civil Aviation, Road Transportation, Railways, 
Water-borne Navigation, Other Transportation
- **Residential** — Residential and Other Sectors
- **Fugitive** — Solid Fuels, Oil and Natural Gas, Fossil Fuel Fires
- **Agriculture** — Liming, Urea Application
- **Other** — Non-Specified, Non-Energy Products from Fuels and 
Solvent Use

This grouping follows standard practice in sectoral emission analysis 
and ensures that each group represents a meaningful and distinct 
economic activity. The aggregation was validated by confirming that 
the sum of all sector group emissions matched the EDGAR country total 
for every year in the study period, with residual differences 
attributable only to floating point precision and not to data loss.

For each sector group and year, two derived variables were calculated. 
First, the sector share, defined as each sector's emissions as a 
proportion of total national emissions, capturing the structural 
composition of India's emission profile. Second, the emission 
intensity, defined as each sector's emissions per unit of real GDP, 
capturing the combined effect of energy efficiency and energy mix 
within each sector.

The three LMDI effects were then calculated year by year using the 
log mean weighting formula, with each effect summed across all seven 
sector groups to produce national level decomposition results.

## The Environmental Kuznets Curve

The Environmental Kuznets Curve (EKC) hypothesis proposes that the 
relationship between per capita income and environmental degradation 
follows an inverted U shape. As income rises from low levels, emissions 
initially increase due to industrialisation. Beyond a certain income 
threshold, structural change toward services, greater demand for 
environmental quality, and access to cleaner technologies cause 
emissions to decline.

To test the EKC hypothesis for India over the period 2000 to 2022, the 
following quadratic regression model is estimated:

$$CO_{2,t} = \alpha + \beta_1 Y_t + \beta_2 Y_t^2 + \beta_3 X_t + \varepsilon_t$$

Where $CO_{2,t}$ is per capita CO2 emissions in year $t$, $Y_t$ is 
real GDP per capita, $Y_t^2$ is the squared term of GDP per capita, 
and $X_t$ represents a vector of control variables including energy 
intensity and renewable energy share. The EKC hypothesis is confirmed 
if $\beta_1 > 0$ and $\beta_2 < 0$, both statistically significant. 
The income turning point is then estimated as:

$$Y^* = -\frac{\beta_1}{2\beta_2}$$

**Data Processing:**

For the EKC regression, India's total CO2 emissions from the EDGAR 
TOTALS BY COUNTRY dataset were divided by population to derive per 
capita emissions, the dependent variable in the model. Real GDP from 
the World Bank World Development Indicators was similarly divided by 
population to produce GDP per capita in constant 2015 USD. A squared 
term of GDP per capita was computed to capture the non-linear inverted 
U relationship. Two control variables were included: the share of 
renewables in total energy consumption and energy consumption per unit 
of GDP, both sourced from the Our World in Data energy dataset. All 
variables were filtered to the study period 2000 to 2022, producing 
23 annual observations for the regression. Population data were 
obtained from the Our World in Data dataset, which sources demographic 
data from the United Nations World Population Prospects.

## Data Sources

All data used in this study are drawn from publicly available 
institutional sources. CO2 and GHG emission data are obtained from the 
EDGAR (Emissions Database for Global Atmospheric Research) database 
maintained by the European Commission Joint Research Centre. GDP and 
GDP per capita data in constant 2015 USD are sourced from the World 
Bank World Development Indicators. Energy consumption data by fuel type 
are obtained from Our World in Data, which compiles and harmonises 
energy statistics from the BP Statistical Review of World Energy. State 
level electricity generation data are obtained from the Central 
Electricity Authority of India and are used in the spatial analysis 
component of this study. 

Throughout this study, emissions are reported in Gigagrams of CO2 
(Gg CO2), the native unit used by the EDGAR database. One Gigagram 
is equivalent to one thousand metric tonnes, or one million kilograms. 
For reference, India's total CO2 emissions in 2022 were approximately 
2,741,000 Gg CO2, equivalent to 2.74 billion metric tonnes. This unit 
is retained throughout to avoid conversion errors and ensure 
reproducibility against the source data.

The study period covers 2000 to 2022, 
representing the most complete and consistent data available across all 
sources at the time of writing.

# Results and Discussion

## Tapio Decoupling Analysis

### Results

The analysis covers the period 2001 to 2022, with 2000 serving as the 
base year for calculating year on year percentage changes and therefore 
excluded from the results.

```{r}
#| label: tbl-tapio
#| tbl-cap: "Tapio Decoupling Index Results for India (2001 to 2022)"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# Load data
panel_data <- read_csv(here::here("data/processed/panel_data.csv"))

# Calculate Tapio index
tapio_data <- panel_data %>%
  arrange(year) %>%
  mutate(
    pct_change_co2 = (co2_total - lag(co2_total)) / lag(co2_total),
    pct_change_gdp = (gdp - lag(gdp)) / gdp,
    tapio_index = pct_change_co2 / pct_change_gdp
  ) %>%
  filter(!is.na(tapio_index)) %>%
  mutate(
    decoupling_state = case_when(
      pct_change_gdp > 0 & tapio_index < 0 ~ "Strong Decoupling",
      pct_change_gdp > 0 & tapio_index >= 0 & tapio_index < 0.8 ~ "Weak Decoupling",
      pct_change_gdp > 0 & tapio_index >= 0.8 & tapio_index < 1.2 ~ "Expansive Coupling",
      pct_change_gdp > 0 & tapio_index >= 1.2 ~ "Expansive Negative Coupling",
      pct_change_gdp < 0 & tapio_index > 0 ~ "Recessive Coupling",
      pct_change_gdp < 0 & tapio_index < 0 ~ "Recessive Decoupling",
      TRUE ~ "Undefined"
    )
  )

# Display table
tapio_table <- tapio_data %>%
  select(year, tapio_index, decoupling_state) %>%
  mutate(tapio_index = round(tapio_index, 3)) %>%
  rename(
    Year = year,
    `Tapio Index` = tapio_index,
    `Decoupling State` = decoupling_state
  )

kable(tapio_table, align = "c")
```

```{r}
#| label: fig-tapio
#| fig-cap: "Tapio Decoupling Index for India (2001 to 2022)"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)

# Load processed panel data
panel_data <- read_csv(here::here("data/processed/panel_data.csv"))

# Calculate Tapio index
tapio_data <- panel_data %>%
  arrange(year) %>%
  mutate(
    pct_change_co2 = (co2_total - lag(co2_total)) / lag(co2_total),
    pct_change_gdp = (gdp - lag(gdp)) / gdp,
    tapio_index = pct_change_co2 / pct_change_gdp
  ) %>%
  filter(!is.na(tapio_index)) %>%
  mutate(
    decoupling_state = case_when(
      pct_change_gdp > 0 & tapio_index < 0 ~ "Strong Decoupling",
      pct_change_gdp > 0 & tapio_index >= 0 & tapio_index < 0.8 ~ "Weak Decoupling",
      pct_change_gdp > 0 & tapio_index >= 0.8 & tapio_index < 1.2 ~ "Expansive Coupling",
      pct_change_gdp > 0 & tapio_index >= 1.2 ~ "Expansive Negative Coupling",
      pct_change_gdp < 0 & tapio_index > 0 ~ "Recessive Coupling",
      pct_change_gdp < 0 & tapio_index < 0 ~ "Recessive Decoupling",
      TRUE ~ "Undefined"
    )
  )

# Define colours
state_colours <- c(
  "Strong Decoupling"           = "#2ecc71",
  "Weak Decoupling"             = "#a8d8a8",
  "Expansive Coupling"          = "#f39c12",
  "Expansive Negative Coupling" = "#e74c3c",
  "Recessive Coupling"          = "#9b59b6"
)

# Plot
ggplot(tapio_data, aes(x = year, y = tapio_index, fill = decoupling_state)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "gray20", linewidth = 0.7) +
  geom_hline(yintercept = 1.2, linetype = "dashed", color = "gray20", linewidth = 0.7) +
  scale_fill_manual(values = state_colours) +
  scale_x_continuous(breaks = seq(2001, 2022, by = 2)) +
  labs(
    title = "Tapio Decoupling Index for India (2001 to 2022)",
    x = "Year",
    y = "Tapio Decoupling Index",
    fill = "Decoupling State",
    caption = "Source: EDGAR, World Bank WDI. Authors own calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.4, "cm"),
    legend.title = element_text(size = 9, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Figure 1 presents the year by year Tapio decoupling index values and 
their corresponding classification for India over the study period 
2001 to 2022. The results reveal a highly inconsistent decoupling 
trajectory, with India oscillating between weak decoupling and 
expansive negative coupling across the two decades studied.

During the early study period from 2001 to 2006, India predominantly 
exhibited weak decoupling, suggesting that GDP growth was outpacing 
emission growth during this phase. However, the period from 2007 to 
2012 saw a marked deterioration, with four of six years classified as 
expansive negative coupling, indicating that emissions were growing 
faster than the economy. This coincides with India's rapid 
industrialisation and significant expansion of coal based power 
generation during this period.

A partial recovery toward weak decoupling was observed between 2013 
and 2016, potentially reflecting early investments in renewable energy 
capacity and improvements in energy efficiency. The single instance of 
strong decoupling occurred in 2019, the only year in the entire study 
period where CO2 emissions declined while GDP continued to grow. This 
finding warrants closer examination in the LMDI decomposition analysis 
to identify its underlying drivers.

The year 2020 recorded recessive coupling, where both GDP and emissions 
fell simultaneously as a consequence of the COVID-19 pandemic induced 
economic contraction. This does not represent genuine environmental 
progress and must be interpreted with caution. The post pandemic 
recovery in 2021 and 2022 returned India to expansive coupling, 
suggesting that emission reductions observed during the pandemic were 
temporary rather than structural.

### Discussion

The Tapio decoupling analysis reveals that India has not achieved 
sustained decoupling of CO2 emissions from GDP growth over the 2001 
to 2022 period. While episodes of weak decoupling were observed, 
particularly in the early 2000s and mid 2010s, these were not 
maintained consistently, and the overall trajectory suggests that 
economic growth in India has continued to drive emission increases 
across most of the study period.

The deterioration toward expansive negative coupling between 2007 and 
2012 is particularly significant. This period coincided with India's 
rapid expansion of coal based power generation to meet rising 
electricity demand from industrialisation and urban growth. The 
findings suggest that energy supply decisions during this decade had 
a lasting impact on India's emission trajectory that subsequent 
efficiency improvements have only partially offset.

The single instance of strong decoupling in 2019 is a noteworthy 
finding. While it represents the only year in the study period where 
emissions fell alongside positive GDP growth, it should be interpreted 
cautiously. A single year of strong decoupling does not constitute a 
structural shift, and the subsequent return to expansive coupling in 
2021 and 2022 confirms that this was not the beginning of a sustained 
transition. The LMDI decomposition analysis in the following section 
will investigate the specific drivers behind this observation.

The 2020 recessive coupling result, driven by the COVID-19 pandemic, 
highlights an important methodological consideration. Emission 
reductions caused by economic contraction are fundamentally different 
from those achieved through structural change or technological 
transition. Including 2020 in trend analysis without this caveat would 
risk overstating India's decoupling progress.

Overall the Tapio results raise significant questions about whether 
India's declining emission intensity, as reported in its NDC progress 
assessments, reflects genuine structural decoupling or is partly 
attributable to the compositional shift of the economy toward less 
energy intensive services. This question is addressed directly in the 
LMDI decomposition that follows.

## LMDI Decomposition Analysis

### Results

The LMDI decomposition separates the total change in India's CO2 
emissions each year into three contributing effects: the activity 
effect, reflecting changes driven by overall GDP growth; the structure 
effect, capturing shifts in the sectoral composition of the economy; 
and the intensity effect, representing changes in emission intensity 
across sectors. The year 2000 serves as the base year and is excluded 
from the decomposition results.

```{r}
#| label: tbl-lmdi
#| tbl-cap: "LMDI Decomposition Results for India (2001 to 2022). All values in Gg CO2."
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(kableExtra)

lmdi_table <- read_csv(here::here("outputs/tables/lmdi_results.csv")) %>%
  mutate(
    year = as.integer(year),
    across(c(activity_effect, structure_effect,
             intensity_effect, actual_change),
           ~ format(round(., 0), big.mark = ","))
  ) %>%
  select(year, activity_effect, structure_effect,
         intensity_effect, actual_change)

kable(lmdi_table, align = "c",
      col.names = c("Year",
                    "Activity Effect (Gg CO2)",
                    "Structure Effect (Gg CO2)",
                    "Intensity Effect (Gg CO2)",
                    "Actual Change (Gg CO2)"))
```

```{r}
#| label: fig-lmdi-overview
#| fig-cap: "LMDI Decomposition Overview with Actual Emission Change (2001 to 2022)"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(patchwork)

# Load data
panel_data <- read_csv(here::here("data/processed/panel_data.csv"))
edgar_sector_long <- read_csv(here::here("data/processed/edgar_sector_long.csv"))

# Log mean function
log_mean <- function(a, b) {
  if (is.na(a) | is.na(b)) return(NA)
  if (a == b) return(a)
  if (a <= 0 | b <= 0) return(0)
  (a - b) / (log(a) - log(b))
}
log_mean_vec <- Vectorize(log_mean)

# Prepare LMDI data
lmdi_data <- edgar_sector_long %>%
  left_join(panel_data %>% select(year, gdp, co2_total), by = "year") %>%
  group_by(year) %>%
  mutate(
    sector_share = co2 / co2_total,
    emission_intensity = co2 / gdp
  ) %>%
  ungroup() %>%
  arrange(sector_group, year)

# Calculate effects
lmdi_results <- lmdi_data %>%
  group_by(sector_group) %>%
  mutate(
    lm_weight = log_mean_vec(co2, lag(co2)),
    activity_effect = lm_weight * log(gdp / lag(gdp)),
    structure_effect = lm_weight * log(sector_share / lag(sector_share)),
    intensity_effect = lm_weight * log(emission_intensity / lag(emission_intensity))
  ) %>%
  ungroup() %>%
  filter(!is.na(activity_effect)) %>%
  group_by(year) %>%
  summarise(
    activity_effect  = sum(activity_effect, na.rm = TRUE),
    structure_effect = sum(structure_effect, na.rm = TRUE),
    intensity_effect = sum(intensity_effect, na.rm = TRUE)
  ) %>%
  left_join(
    panel_data %>%
      arrange(year) %>%
      mutate(actual_change = co2_total - lag(co2_total)) %>%
      select(year, actual_change),
    by = "year"
  )

# Reshape for overview plot
lmdi_long <- lmdi_results %>%
  select(year, activity_effect, structure_effect, intensity_effect) %>%
  pivot_longer(cols = c(activity_effect, structure_effect, intensity_effect),
               names_to = "effect",
               values_to = "value") %>%
  mutate(effect = case_when(
    effect == "activity_effect"  ~ "Activity Effect",
    effect == "structure_effect" ~ "Structure Effect",
    effect == "intensity_effect" ~ "Intensity Effect"
  ))

effect_colours <- c(
  "Activity Effect"  = "#e74c3c",
  "Structure Effect" = "#f39c12",
  "Intensity Effect" = "#2ecc71"
)

# Overview plot
ggplot() +
  geom_col(data = lmdi_long,
           aes(x = year, y = value, fill = effect),
           position = "stack") +
  geom_line(data = lmdi_results,
            aes(x = year, y = actual_change),
            color = "black", linewidth = 1) +
  geom_point(data = lmdi_results,
             aes(x = year, y = actual_change),
             color = "black", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = effect_colours) +
  scale_x_continuous(breaks = seq(2001, 2022, by = 2)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "LMDI Decomposition of CO2 Emission Changes in India (2001 to 2022)",
    subtitle = "Black line shows actual emission change",
    x = "Year",
    y = "Emission Change (Gg CO2)",
    fill = "Effect",
    caption = "Source: EDGAR, World Bank WDI. Authors own calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    legend.position = "right",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
#| label: fig-lmdi-panels
#| fig-cap: "LMDI Decomposition Effects for India by Component (2001 to 2022)"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 14
#| fig-height: 5

p_activity <- ggplot(lmdi_results, aes(x = year, y = activity_effect)) +
  geom_col(fill = "#e74c3c") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous(breaks = seq(2001, 2022, by = 3)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Activity Effect", x = "Year", y = "Emission Change (Gg CO2)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11),
        axis.text.x = element_text(angle = 45, hjust = 1))

p_structure <- ggplot(lmdi_results, aes(x = year, y = structure_effect)) +
  geom_col(fill = "#f39c12") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous(breaks = seq(2001, 2022, by = 3)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Structure Effect", x = "Year", y = "Emission Change (Gg CO2)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11),
        axis.text.x = element_text(angle = 45, hjust = 1))

p_intensity <- ggplot(lmdi_results, aes(x = year, y = intensity_effect)) +
  geom_col(fill = "#2ecc71") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_continuous(breaks = seq(2001, 2022, by = 3)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Intensity Effect", x = "Year", y = "Emission Change (Gg CO2)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 11),
        axis.text.x = element_text(angle = 45, hjust = 1))

p_activity + p_structure + p_intensity +
  plot_annotation(
    title = "LMDI Decomposition Effects for India (2001 to 2022)",
    caption = "Source: EDGAR, World Bank WDI. Authors own calculations.",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )
```

The decomposition results reveal three distinct patterns across the 
study period. The activity effect is consistently positive throughout 
2001 to 2022, reflecting India's sustained economic growth and its 
upward pressure on emissions in every year except 2020. The intensity 
effect is the dominant counterforce, showing large negative values in 
years of strong efficiency improvement and renewable energy expansion, 
particularly 2015 and 2016, but turning positive during periods of 
coal-intensive industrial expansion such as 2007 to 2012. The 
structure effect is negligible across the entire study period, with 
values close to zero in every year, indicating that shifts in 
sectoral composition contributed minimally to emission changes.

### Discussion

The LMDI decomposition results provide critical insight into the 
mechanisms underlying India's emission trajectory and directly address 
the central research question of whether observed decoupling reflects 
genuine structural change.

The most significant finding is the near-zero structure effect across 
all years of the study period. This result challenges the widely held 
narrative that India's transition toward a service-dominated economy 
has been a meaningful driver of emission reductions. While India's 
services sector has grown substantially as a share of GDP during 2000 
to 2022, the sectoral composition of emissions has remained largely 
unchanged, suggesting that emission-intensive sectors such as power 
generation and manufacturing have maintained their relative share of 
total emissions even as the broader economy evolved.

The intensity effect emerges as the primary mechanism through which 
India has achieved periods of emission reduction or moderation. The 
large negative intensity values observed in 2015 and 2016 coincide 
with India's accelerated deployment of renewable energy capacity 
following the launch of the National Solar Mission targets in 2010 
and subsequent policy push. This suggests that technology and policy 
driven changes in the energy mix, rather than structural economic 
change, have been responsible for India's improvements in emission 
intensity.

Conversely, the positive intensity values observed between 2007 and 
2012 indicate a period when coal-based capacity additions outpaced 
efficiency gains, pushing emission intensity upward. This finding 
aligns with historical records of rapid coal power plant construction 
during this period to address India's chronic electricity deficit.

The dominance of the activity effect throughout the study period 
confirms that economic growth remains the primary driver of rising 
absolute emissions in India. Without sustained and large negative 
intensity effects, India's emission trajectory would have been 
significantly steeper. This underscores the critical importance of 
continued investment in renewable energy and energy efficiency to 
counterbalance the emission pressure of economic growth, particularly 
as India targets higher GDP growth rates in the coming decade.

## EKC Regression Analysis

### Results

The Environmental Kuznets Curve hypothesis is tested using a quadratic 
regression of per capita CO2 emissions on GDP per capita, its squared 
term, and two control variables covering the period 2000 to 2022. The 
estimated regression equation is:

$$\widehat{CO_{2,pc}} = -0.00336 + 0.00000255 \cdot Y - 0.000000000408 \cdot Y^2 - 0.0000385 \cdot REN + 0.00215 \cdot EI$$

Where $Y$ is GDP per capita in constant 2015 USD, $REN$ is the share 
of renewables in total energy consumption, and $EI$ is energy 
consumption per unit of GDP.

```{r}
#| label: tbl-ekc
#| tbl-cap: "EKC Regression Results for India (2000 to 2022). Note: *** p < 0.001, ** p < 0.01, * p < 0.05. Coefficients and standard errors are reported in their original scale."
#| echo: false
#| warning: false
#| message: false

library(knitr)
library(kableExtra)

ekc_reg_table <- data.frame(
  Variable = c("Intercept", 
               "GDP per Capita (USD)", 
               "GDP per Capita Squared (USD²)",
               "Renewables Share (% of total energy)", 
               "Energy Intensity (energy per GDP)"),
  Coefficient = c("-3.358 \u00d7 10\u207b\u00b3", 
                  "2.548 \u00d7 10\u207b\u2076", 
                  "-4.077 \u00d7 10\u207b\u00b9\u2070",
                  "-3.847 \u00d7 10\u207b\u2075", 
                  "2.154 \u00d7 10\u207b\u00b3"),
  Std_Error = c("6.837 \u00d7 10\u207b\u2074", 
                "1.986 \u00d7 10\u207b\u2077",
                "5.419 \u00d7 10\u207b\u00b9\u00b9",
                "1.111 \u00d7 10\u207b\u2075", 
                "4.438 \u00d7 10\u207b\u2074"),
  t_value = c("-4.912", "12.835", "-7.522", "-3.462", "4.853"),
  p_value = c("< 0.001 ***", "< 0.001 ***", "< 0.001 ***",
              "0.003 **", "< 0.001 ***")
)

kable(ekc_reg_table, align = "c",
      col.names = c("Variable", "Coefficient",
                    "Std. Error", "t-value", "p-value"))
```

The model demonstrates an exceptionally strong fit, with an R-squared 
of 0.9915 and an adjusted R-squared of 0.9896, indicating that GDP 
per capita, renewable energy share, and energy intensity together 
explain 99.15% of the variation in India's per capita CO2 emissions 
over the study period. The overall model is highly statistically 
significant, with an F-statistic of 525.3 and a p-value below 0.001.

```{r}
#| label: fig-ekc
#| fig-cap: "Environmental Kuznets Curve for India (2000 to 2022) with Projected Turning Point"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(here)

panel_data <- read_csv(here::here("data/processed/panel_data.csv"))

ekc_data <- panel_data %>%
  mutate(
    co2_pc = co2_total / population,
    gdp_pc = gdp / population,
    gdp_pc_sq = gdp_pc^2
  ) %>%
  select(year, co2_pc, gdp_pc, gdp_pc_sq,
         renewables_share_energy, energy_per_gdp)

ekc_model <- lm(co2_pc ~ gdp_pc + gdp_pc_sq +
                  renewables_share_energy + energy_per_gdp,
                data = ekc_data)

beta1 <- coef(ekc_model)["gdp_pc"]
beta2 <- coef(ekc_model)["gdp_pc_sq"]
turning_point <- -beta1 / (2 * beta2)

recent_growth <- ekc_data %>%
  filter(year >= 2015) %>%
  summarise(avg_growth = mean((gdp_pc - lag(gdp_pc)) / lag(gdp_pc), 
                               na.rm = TRUE)) %>%
  pull(avg_growth)

last_gdp_pc <- ekc_data %>% filter(year == 2022) %>% pull(gdp_pc)

projected <- data.frame(year = 2023:2045) %>%
  mutate(gdp_pc = last_gdp_pc * (1 + recent_growth)^(year - 2022)) %>%
  filter(gdp_pc <= turning_point * 1.1)

turning_year <- projected %>%
  filter(gdp_pc >= turning_point) %>%
  slice(1) %>%
  pull(year)

gdp_range <- data.frame(
  gdp_pc = seq(min(ekc_data$gdp_pc), turning_point * 1.05, length.out = 300),
  renewables_share_energy = mean(ekc_data$renewables_share_energy, na.rm = TRUE),
  energy_per_gdp = mean(ekc_data$energy_per_gdp, na.rm = TRUE)
) %>%
  mutate(gdp_pc_sq = gdp_pc^2)

gdp_range$co2_pc_fitted <- predict(ekc_model, newdata = gdp_range)

ggplot() +
  geom_line(data = gdp_range %>% filter(gdp_pc <= last_gdp_pc),
            aes(x = gdp_pc, y = co2_pc_fitted),
            color = "#2c3e50", linewidth = 1.2) +
  geom_line(data = gdp_range %>% filter(gdp_pc >= last_gdp_pc),
            aes(x = gdp_pc, y = co2_pc_fitted),
            color = "#2c3e50", linewidth = 1.2, linetype = "dashed") +
  geom_point(data = ekc_data,
             aes(x = gdp_pc, y = co2_pc),
             color = "#e74c3c", size = 2.5) +
  geom_segment(data = ekc_data %>%
               filter(year %in% c(2000, 2005, 2010, 2015, 2019, 2022)) %>%
               mutate(
                 nudge_y = case_when(
                   year %in% c(2000, 2005, 2010, 2022) ~ 0.00014,
                   year %in% c(2015, 2019) ~ -0.00014
                 )
               ),
             aes(x = gdp_pc, y = co2_pc + nudge_y * 0.85,
                 xend = gdp_pc, yend = co2_pc + nudge_y * 0.15),
             color = "gray30", linewidth = 0.5,
             arrow = arrow(length = unit(0.008, "npc"), type = "closed")) +
geom_text(data = ekc_data %>%
            filter(year %in% c(2000, 2005, 2010, 2015, 2019, 2022)) %>%
            mutate(
              nudge_y = case_when(
                year %in% c(2000, 2005, 2010, 2022) ~ 0.00014,
                year %in% c(2015, 2019) ~ -0.00014
              )
            ),
          aes(x = gdp_pc, y = co2_pc + nudge_y, label = year),
          size = 3, color = "#e74c3c") +
  geom_vline(xintercept = turning_point,
             linetype = "dashed", color = "#f39c12", linewidth = 0.8) +
  annotate("text", x = turning_point - 80,
           y = max(gdp_range$co2_pc_fitted) * 0.85,
           label = paste0("Turning point\nUSD ", round(turning_point, 0),
                          "\n(~", turning_year, ")"),
           hjust = 1, size = 3.2, color = "#f39c12") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Environmental Kuznets Curve for India (2000 to 2022) with Projection",
    subtitle = "Solid line: observed period. Dashed line: projected trajectory. Red points: actual observations.",
    x = "GDP per Capita (constant 2015 USD)",
    y = "CO2 Emissions per Capita (Gg CO2)",
    caption = "Source: EDGAR, World Bank WDI. Authors own calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray40")
  )
```

```{r}
#| label: fig-ekc-fit
#| fig-cap: "EKC Model Fit: Actual vs Fitted CO2 Per Capita for India (2000 to 2022)"
#| echo: false
#| warning: false
#| message: false

ekc_data_fitted <- ekc_data %>%
  mutate(fitted = fitted(ekc_model))

ggplot(ekc_data_fitted, aes(x = year)) +
  geom_line(aes(y = co2_pc, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = fitted, color = "Fitted"), linewidth = 1, linetype = "dashed") +
  geom_point(aes(y = co2_pc, color = "Actual"), size = 2) +
  scale_color_manual(values = c("Actual" = "#e74c3c", "Fitted" = "#2c3e50")) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = seq(2000, 2022, by = 2)) +
  labs(
    title = "EKC Model Fit: Actual vs Fitted CO2 Per Capita for India (2000 to 2022)",
    subtitle = "R-squared = 0.991, Adjusted R-squared = 0.990, F-statistic p-value < 0.001",
    x = "Year",
    y = "CO2 Emissions per Capita (Gg CO2)",
    color = "",
    caption = "Source: EDGAR, World Bank WDI. Authors own calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Discussion

The EKC regression results confirm the existence of an inverted U 
shaped relationship between per capita income and CO2 emissions for 
India over the 2000 to 2022 period. The positive and statistically 
significant coefficient on GDP per capita and the negative and 
statistically significant coefficient on its squared term together 
satisfy the mathematical conditions for the EKC hypothesis, consistent 
with findings by Sinha and Shahbaz (2018) and Gopakumar et al. (2022) 
for earlier study periods.

A notable feature of the regression results is the systematic shift 
in the relationship between actual and fitted values around 2015. 
Prior to 2015, India's actual per capita CO2 emissions consistently 
exceeded the model's fitted values, indicating that emissions were 
higher than what GDP per capita alone would predict. From 2015 
onwards, actual emissions fell below the fitted values, suggesting 
that India became cleaner than the income driven EKC trajectory 
would anticipate. This structural break coincides with two major 
policy interventions. First, the Jawaharlal Nehru National Solar 
Mission, launched in January 2010, began deploying significant 
solar capacity at scale during its Phase II from 2014 to 2017. 
Second, in 2015 India's Ministry of New and Renewable Energy adopted 
an ambitious revised target of 175 GW of total renewable energy 
capacity by 2022, comprising 100 GW of solar, 60 GW of wind, and 
15 GW from biomass and small hydro. The acceleration of renewable 
deployment following these policy commitments appears to have driven 
actual emissions below the model baseline, providing evidence that 
deliberate policy intervention, rather than income growth alone, 
was responsible for India's emission improvements after 2015. This 
finding reinforces the LMDI result that the intensity effect was 
strongly negative in 2015 and 2016, and together these two analyses 
point to the same conclusion: India's post 2015 emission moderation 
was policy driven, not structurally determined.

The estimated income turning point of approximately USD 3,126 per 
capita in constant 2015 USD is particularly significant. India's GDP 
per capita stood at approximately USD 2,098 in 2022, meaning India 
remains on the rising left side of the inverted U curve and has not 
yet reached its emission peak. Based on India's average GDP per capita 
growth rate of approximately 4.5% per year observed between 2015 and 
2022, the model projects that India will reach its EKC turning point 
around 2032, assuming current growth and energy policy trajectories 
are maintained.

This projected turning point carries important policy implications. 
First, it coincides closely with India's NDC commitment year of 2030, 
suggesting that achieving the NDC target of reducing emission intensity 
by 45% relative to 2005 levels may require active policy intervention 
rather than relying on income driven structural change alone. Second, 
the significant negative coefficient on renewable energy share 
indicates that accelerating the renewable energy transition could bring 
the turning point forward, potentially before 2032. Conversely, any 
slowdown in renewable deployment or acceleration of fossil fuel 
consumption could push the turning point further into the future.

The exceptionally high R-squared of 0.9915 confirms that the three 
variable model captures India's emission dynamics with very high 
precision. However this should be interpreted with appropriate caution 
given the relatively small sample size of 23 annual observations. The 
model is best understood as a characterisation of India's emission 
trajectory over the specific study period rather than a universal 
predictive tool.

It is also important to note that the EKC turning point estimated here 
is higher than the USD 2,938 per capita estimate reported by Sinha and 
Shahbaz (2018) for an earlier period ending in 2014. This upward 
revision may reflect the continued expansion of coal based capacity 
during 2015 to 2022 that has sustained emission pressure at higher 
income levels than previously anticipated, and underscores the 
importance of updating EKC estimates as new data become available.